\Large Taking uncertainty seriously  \newline \normalsize A  Bayesian approach

### 2022-03-08 presentations/presentationBoston/tryOut2.Rmd
presentation Boston added cosine etc.

---
title: '\Large Taking uncertainty seriously  \newline \normalsize A  Bayesian approach
  to  word embedding bias estimation '
author: Alicja Dobrzeniecka \& Rafal  Urbaniak \footnotesize \newline (LoPSE research
  group, University of  Gdansk)
date: "Boston, April Fools' Day"
output:
  beamer_presentation:
    theme: Rafal_beamerSly1
    keep_tex: yes
    slide_level: 2
  slidy_presentation: default
fontsize: 10pt
classoption: x11names, dvipsnames, bibspacing,natbib, table
urlcolor: blue
bibliography: ../../references/cosineReferences1.bib
csl: ../../references/apa-6th-edition.csl
link-citations: yes
---

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(ggExtra)
library(ggthemes)
library(ggforce)
library(latex2exp)
th <- theme_tufte(base_size = 20)
options(kableExtra.latex.load_packages = FALSE)
```

## Presentation plan

- Bias in word embeddings
- WEAT and MAC methods
- Methodological problems
- Limitations of pre-averaging in bias detection methods
- Accounting for uncertainty with Bayesian approach

@@

word | 1 | 2 | 3 | 4 | ... 
|---|---|---|---|---|---
woman | 0.456 | 0.267 | 0.675 | 0.131| ...
man | 0.451 | 0.897 | 0.472 | 0.088| ...

How is this supposed to capture semantic relations?

@@

### General idea

Similarity in vector direction.

## Cosine similarity

```{r cosine1,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
a <- c(1,2)
b <- c(3,2.5)

ggplot()+
  xlim(c(0,4.5))+ylim(c(0,4.5))+
  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
               arrow = arrow(length = unit(0.02, "npc")))+
  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
  theme_tufte(base_size = 24)

```

\begin{block}{Vectors}
\begin{align*}
a  & = [1,2]\\
b  &= [3,2]
\end{align*}

\end{block}
\pause

\begin{block}{Dot product}

\begin{align*}
a \cdot b & = a_1 b_1 + a_2 b_2\\
a \cdot a & = a_1^2 + a_2 ^ 2 \\
\lVert a\rVert & = \sqrt(a \cdot a)
\end{align*}

## Cosine similarity

```{r cosine2,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
a <- c(1,2)
b <- c(3,2.5)
bma <- b - a

ggplot()+
  xlim(c(0,4.5))+ylim(c(0,4.5))+
  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
               arrow = arrow(length = unit(0.02, "npc")))+
  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
  theme_tufte(base_size = 24)+
  geom_segment(aes(x = 0, y = 0, xend = bma[1], yend = bma[2]),
      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
  annotate(geom = "text", label  = "b - a = [2,.5]", x = 2.2, y = .9, size = 7)+
  theme_tufte(base_size = 24)

```

\begin{block}{Vectors}

\begin{align*}
a  & = [1,2]\\
b  &= [4,4]
\end{align*}

\begin{block}{Dot product}

\begin{align*}
a \cdot b & = a_1 b_1 + a_2 b_2\\
a \cdot a & = a_1^2 + a_2 ^ 2 \\
\lVert a\rVert & = \sqrt(a \cdot a)
\end{align*}

\begin{block}{Vector difference}

\begin{align*}
b - a & = [b_1- a_1, b_2 - a_2 ]
\end{align*}

## Cosine similarity

```{r cosine3,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
a <- c(1,2)
b <- c(3,2.5)

ggplot()+
  xlim(c(0,4.5))+ylim(c(0,4.5))+
  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
               arrow = arrow(length = unit(0.02, "npc")))+
  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
  theme_tufte(base_size = 24)+
  geom_segment(aes(x = a[1], y = a[2], xend = b[1], yend = b[2]),
      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
  annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
  theme_tufte(base_size = 24)

```

\begin{block}{Vectors}

\begin{align*}
a  & = [1,2]\\
b  &= [4,4]
\end{align*}

\begin{block}{Dot product}

\begin{align*}
a \cdot b & = a_1 b_1 + a_2 b_2\\
a \cdot a & = a_1^2 + a_2 ^ 2 \\
\lVert a\rVert & = \sqrt(a \cdot a)
\end{align*}

\begin{block}{Vector difference}

\begin{align*}
b - a & = [b_1- a_1, b_2 - a_2 ]
\end{align*}

## Cosine similarity

```{r cosine4,echo=FALSE,eval=TRUE,fig.align = "center", cache=TRUE, fig.show = "hold", out.width = "100%",   message = FALSE, warning = FALSE, results = FALSE}
a <- c(1,2)
b <- c(3,2.5)
#atan2(a[1] , a[2])
#atan2(b[1] , b[2])

ggplot()+
  xlim(c(0,4.5))+ylim(c(0,4.5))+
  geom_segment(aes(x = 0, y = 0, xend = a[1], yend = a[2]),  
               arrow = arrow(length = unit(0.02, "npc")))+
  annotate(geom = "text", label  = "a = [1,2]", x = 1.2, y = 2.4, size = 7)+
  geom_segment(aes(x = 0, y = 0, xend = b[1], yend = b[2]), arrow = arrow(length = unit(0.02, "npc")))+   annotate(geom = "text", label  = "b = [3,2.5]", x = 3.2, y = 2.9, size = 7)+
  theme_tufte(base_size = 24)+
  geom_segment(aes(x = a[1], y = a[2], xend = b[1], yend = b[2]),
      arrow = arrow(length = unit(0.02, "npc")), color = "grey")+ 
  annotate(geom = "text", label  = "b - a = [2,.5] (shifted)", x = 2, y = 2.7, size = 7)+
  theme_tufte(base_size = 24)+
  annotate(geom = "text", label  = expression(theta), x = .6, y = .78, size = 7)+
  geom_arc(aes(x0 = 0, y0 = 0, r = 1.3, 
               start = .463, end = .876))

```

\begin{block}{Angle}

\begin{align*}
\lVert b - a \rVert^2 &  = \lVert  b \rVert^2 + \lVert a \rVert^2 - 
  2 \lVert b \rVert   \lvert  a \rVert  \cos \theta  \\
b \cdot a  & = \lVert b\rVert \lVert  a \rVert \cos \theta \\ 
\cos \theta & = \frac{b \cdot a}{\lVert b\rVert  \lVert  a \rVert }
\end{align*}

\begin{block}{Orthogonality}
\begin{align*}
\cos ( 90^{\circ}) & = 0 \\ 
\frac{b \cdot a}{\lVert  b\rVert \lVert a \rVert} & = 0\\ 
 b \cdot a & = 0  
\end{align*}
\end{block}

## Cosine similarity \& distance

<!-- - Dot product for normalized vectors -->

<!-- often used in discussion of bias -->

\vspace{-4mm}

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\lVert  A \rVert \,\lVert B \rVert}
\\
\tag{Distance}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)
\end{align}

-  Geometric interpretation: direction (not length)

-  $\mathsf{cosineDistance}\in (0, 2)$

- Naive interpretation: proximity corresponds to semantic similarity

<!-- - Note cosine distance fails to meet triangle inequality -->

## Cosine-based measures of bias

### The worry

Word embeddings can learn implicit harmful biases

### The basic intuition

Stereotypically connected words are cosine-close

## Cosine-based measures of bias

### A visual example

- "feminine" occupations: "homemaker", "nurse", "receptionist", "librarian", etc.

- "masculine" occupations:  "maestro",  "captain", "architect", "boss", etc.

\normalsize

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "60%"}
genderProjections <-  read.csv("datasets/genderProjections.csv")[,-1]
colnames(genderProjections)[4] <- "Stereotype" 
levels(genderProjections$Stereotype) <- c("male occupation", "female occupation")
ggplot(genderProjections, aes(x=he,y=she, color = Stereotype, shape = Stereotype))+geom_point(size = 6, alpha = 0.75)+theme_tufte(base_size = 27)+labs(title ="GloVe  on Wikipedia 2014 and Gigaword 5th ed.")+scale_color_manual(values =c("orangered4","chartreuse4"))+ 
  theme(legend.position = c(0.8,.15)) +
    scale_shape_manual(values=c(15, 19)) 
```
\normalsize

<!-- ## Cosine-based measures of bias -->

<!-- ### Example: direct bias -->

<!-- - The gender bias of a word $w$ is its projection on the gender direction $\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})$ -->

<!-- <!-- (the gender direction is the top principal compontent of ten gender pair difference vectors).  -->

<!-- -  Given the (ideally) gender neutral words $N$ and the gender direction $g$ the direct gender bias is: -->
<!-- <!-- ($c$ is a parameter determining how strict we want to be): -->

<!-- \vspace{-2mm} -->

<!-- \begin{align} -->
<!-- \mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert } -->
<!-- \end{align} -->

<!-- \footnotesize  -->

<!-- [@Bolukbasi2016Man] -->

## Cosine-based measures of bias

### Example: Word Embedding Association Test (WEAT)

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "55%"}
words <- c("man", "he", "boss", "architect","nurse","receptionist","woman","she")
x <- c(.1,.1,.5,.5,.7,.7,1.2,1.2)
y <- c(.4,.3,.3,.2,.5,.4,.4,.3)
weat <- data.frame(words,x,y)

ggplot(weat, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
                                                                 nudge_y = 0.018, size = 7 )+
    geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.2), , linetype =2, arrow=arrow(), size=.05, color="blue") +
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.5, yend=.3), linetype =2, arrow=arrow(), size=.05, color="blue")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+
  geom_segment(data=weat, mapping=aes(x=.1, y=.3, xend=.7, yend=.5), linetype =2, arrow=arrow(), size=.05, color="orangered")+
    annotate("text", x = c(.32,.32,.32,.32), y = c(.23,.29,.35,.39), 
             label = c(".7", ".6",".1", ".2") , color="black", 
             size=7,  fontface="bold")+
    annotate("text", x = c(.1,.5,.7,1.2), y = c(.55,.55,.55,.55), 
           label = c("X", "A","B", "Y") , color=c("black", "blue","orangered","black"), 
           size=10,  fontface="bold")+xlim(0,1.3)+
  theme_void(base_size = 25)
```
\normalsize

- $s_1 = s(he,A,B) =  \frac{.6+.7}{2}  - \frac{.2+.1}{2} = .65-.15= .5$

- $s_2 = s(man,A,B) = .3$, \linebreak  $s_3 = s(woman,A,B) = -.6, s_4 = s(she, A, B) = -.3$

\vspace{-4mm}

\normalsize 
\begin{align*}
\mathsf{WEAT}(A,B) & = \frac{\nicefrac{(s_1+s_2)}{2} - \nicefrac{(s_3+s_4)}{2}}{sd(\{s_1,s_2,s_3,s_4\})} \approx 1.93
\end{align*}

## Cosine-based measures of bias

### Example: Word Embedding Association Test (WEAT)

\begin{align*}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\\
WEAT(A,B) & = \frac{
\mu\left(\{s(x,A,B)\}_{x\in X}\right) -\mu\left(\{s(y,A,B)\}_{y\in Y}\right) 
}{
\sigma\left(\{s(w,A,B)\}_{w\in X\cup Y}\right)
}
\end{align*}

-  $t$ is a term,  $A, B$ are sets of stereotype attribute words, $X$, $Y$ are protected group words

- For instance, $X$ might be a set of male names, $Y$ a set of female names, $A$ might contain stereotypically male-related career words, and $B$ stereotypically female-related family words

- $s$-values are used as datapoints in statistical significance tests

\vspace{2mm}
[@Caliskan2017semanticsBiases] with extensions in [@Lauscher2019multidimensional] and applications in [@Garg2018years]

## Cosine-based measures of bias

### Our main target: Mean Average Cosine Similarity (MAC)

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "50%"}
words <- c("jew", "muslim", "christian", "terrorist","dirty","conservative","familiar","greedy","cheap")
x <- c(.1,.1,.1,.4,.4,.7,.7,1.1,1.1)
y <- c(.4,.3,.2,.1,.2,.3,.4,.1,.2)
mac <- data.frame(words,x)

ggplot(mac, aes(x=x,y=y, label = words))+geom_point()+geom_text(check_overlap = TRUE,
                                                                 nudge_y = 0.02, size = 9 )+
  geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.1), , linetype =2, arrow=arrow(), size=.03, color="blue")+geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.4, yend=.2), , linetype =4, arrow=arrow(), size=.0, color="blue") + geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.3), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=.7, yend=.4), linetype =2, arrow=arrow(), size=.05, color="orangered")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.1), linetype =3, arrow=arrow(), size=.05, color="chartreuse4")+ geom_segment(data=mac, mapping=aes(x=.1, y=.3, xend=1.1, yend=.2), linetype =2, arrow=arrow(), size=.05, color="chartreuse4")+ annotate("text", x = c(.1,.4,.7,1.2), y = c(.45,.45,.45,.45),  label = c("T", "A[1]","A[2]", "A[3]") ,parse = TRUE, color=c("black", "blue","orangered","chartreuse4"),            size=10,  fontface="bold")+xlim(0,1.3)+theme_void(base_size = 25)

```
\normalsize

\vspace{-4mm}

\footnotesize 
\begin{align*}
 s_1 & = s(muslim,A_1) = \frac{cos(muslim,dirty)+cos(muslim,terrorist)}{2}\\
s_2 & = s(muslim,A_2) = \frac{cos(muslim,familiar)+cos(muslim,conservative)}{2}\\
& \vdots \end{align*}

\vspace{-6mm}

\normalsize 
\begin{align*}
 MAC(T,A) & = \mathsf{mean}(\{s_i \vert i \in 1, \dots, k\})
\end{align*}

## Cosine-based measures of bias

### Our main target: Mean Average Cosine Similarity (MAC)

<!-- - more general, multi-class setting -->

\begin{align*}
S(t_i, A_j) & = \frac{1}{\vert A_j\vert}\sum_{a\in A_j}\mathsf{cos}(t,a) \\
MAC(T,A) & = \frac{1}{\vert T \vert \,\vert A\vert}\sum_{t_i \in T }\sum_{A_j \in A} S(t_i,A_j)
\end{align*}

- $T = \{t_1, \dots, t_k\}$ is a class of protected words

- each $A_j\in A$ is a set of attributes stereotypically associated with a protected word

<!-- - For each protected word $T$ and each attribute class, they first take the mean for this protected word and all attributes in a given attribute class, and then take the mean of thus obtained means for all the protected words.  -->

- The t-tests they employ are run on average cosines used to calculate MAC

\vspace{2mm}

[@Manzini2019blackToCriminal]

## Cosine-based measures of bias

### Our main target: Mean Average Cosine Similarity (MAC)

\footnotesize 
```{r religionTableHeadEarly,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion <- read.csv("datasets/religionReddit.csv")[-1]
colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
                        "cosineDistance","cosineSimilarity","connection")
religion$wordClass <- as.factor(religion$wordClass)
levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
set.seed(127)
religionSmall <- religionSmall[sample(1:nrow(religionSmall),6),-c(3,6)]
rownames(religionSmall) <- c()
religionSmall  %>%  kable(format = "latex",booktabs=T,
                      linesep = "",  escape = FALSE, 
                      caption = "A few rows from the religion dataset") %>%
                      kable_styling(latex_options=c("scale_down"))
```
\normalsize

## Cosine-based measures of bias

### General challenges

- Gender-direction: insufficient indicator of
    bias  \footnotesize  [@Gonen2019lipstick]

\normalsize

- Use of analogies: unreliable \footnotesize  [@Nissim2020fair] \normalsize

- High sensitivity to irrelevant factors \footnotesize  [@zhang2020robustness] \normalsize

## Some methodological problems

###  Word list choice is unprincipled

- We run with it for comparison.

### No design considerations to sample size

- ETHAYARAJH IS YOUR criticizes WEAT uses Bernstein bounds and argues that we would need a bias specific dataset of size at least 11903 to claim that the system is biased (three times larger than WinoBias).

- We show progress can be made with more sensitive Bayesian methods.

### The form of the definition is suspicious

ETHAYARAJH UNDERSTANDING shows if there are two target words only WEAT is always maximal in one direction.

- We show the problem runs deeper and stems from pre-averaging, and we statistically gauge the uncertainty that arises from raw sample sizes.

## Some methodological problems

### No word class distinction and no control group

We make the subclasses clear, add human neutral predicates and neutral predicates for control. We used L2-Reddit corpus and GoogleNews (we present the results for Reddit for brevity).

\footnotesize 
```{r religionTableHeadLate,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
religion <- read.csv("datasets/religionReddit.csv")[-1]
colnames(religion) <- c("protectedWord","wordToCompare","wordClass",
                        "cosineDistance","cosineSimilarity","connection")
religion$wordClass <- as.factor(religion$wordClass)
levels(religion$wordClass) <- c("christian","human","jewish","muslim","neutral")
religionSmall <- religion[religion$wordClass != "neutral" & religion$wordClass != "human" ,]
religionControl <- religion[religion$wordClass == "neutral" | religion$wordClass == "human" ,]
set.seed(123154196)
religionDisplayA <- religionSmall[sample(1:nrow(religionSmall),4),]
religionDisplayB <- religionControl[sample(1:nrow(religionControl),4),]
religionDisplay <- rbind(religionDisplayA,religionDisplayB)
religionDisplay$cosineDistance <- round(religionDisplay$cosineDistance,3)
religionDisplay$cosineSimilarity <- round(religionDisplay$cosineSimilarity,3)
rownames(religionDisplay) <- c()
religionDisplay  %>%  kable(format = "latex",booktabs=T,
                      linesep = "",  escape = FALSE, 
                      caption = "Rows from extended religion dataset.") %>%
                      kable_styling(latex_options=c("scale_down"))
```
\normalsize

## Some methodological problems

### Outliers and surprisingly dissimilar words

We study those by  visualizations and uncertainty estimates.

## Some methodological problems

### Distances for  "muslim"

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%",message =FALSE, warning = FALSE}
library(tidyverse)
muslimWords <- c("imam","islam","mosque","muslim","quran")
muslim <- religion %>% filter(protectedWord %in% muslimWords)
muslimClass <- muslim %>% filter(protectedWord == "muslim")
neutralSample <- sample_n(filter(muslimClass,connection == "none"), 5)
humanSample <- sample_n(filter(muslimClass,connection == "human"), 5)
muslimVis <- muslimClass %>% filter(connection != "none" & connection !="human")
muslimVis <- rbind(muslimVis,neutralSample,humanSample)
#we plug in our visualisation script
source("../../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(muslimVis,"muslim")+theme_void(base_size = 20)
```

## Some methodological problems

### Distances for  "priest"

```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}
library(tidyverse)
priestClass <- religion %>% filter(protectedWord == "priest")
neutralSample <- sample_n(filter(priestClass,connection == "none"), 5)
humanSample <- sample_n(filter(priestClass,connection == "human"), 5)
priestVis <- priestClass %>% filter(connection != "none" & connection !="human")
priestVis <- rbind(priestVis,neutralSample,humanSample)

#we plug in our visualisation script
source("../../functions/visualisationTools.R")
#two arguments: dataset and protected word
visualiseProtected(priestVis,"priest")+theme_void(base_size = 20)
```

## Some methodological problems

### No principled interpretation

Religion Debiasing  | MAC (distance)
------------- | -------------
Biased        | 0.859
Hard Debiased | 0.934
Soft Debiased ($\lambda$ = 0.2) | 0.894

- What values are sufficient for the presence of bias and what differences are sign of real improvement?

- Low $p$-values are not high effect indicators!

- We compare HPDIs.

<!--  What may attract attention is the fact that the value of cosine distance in "Biased" category is already quite high even before debiasing. High cosine distance indicates low cosine similarity between values. One could think that average cosine similarity equal to approximately 0.141 is not significant enough to consider it as bias. However the authors aim to mitigate "biases" in vectors with such great distance to make it even larger. Methodologically the question is, on what basis is this small similarity still considered as a proof of the presence of bias, and whether these small changes are meaningful. This is in general the problem of scale and the lack of universal intervals. In contrast, statistical intervals will help us decide whether a given cosine similarity is high enough to consider the words to be more similar than if we chose them at random. We will use highest posterior density intervals, in line with Bayesian methodology.  -->

## The problem with pre-averaging

### Key conceptual issues

- It throws away information about sample sizes
- It ignores variation in the raw data, which leads to false confidence

### Our simulations

Suppose all similarities for two classes are randomly drawn from the same distribution, $\mathsf{Normal}(\mu = 0, \sigma = .08)$, you still can get a really high WEAT!

## The problem with pre-averaging

### Simple case: two pws, four terms

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
set.seed(123)
t1 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
t2 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
t3 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
t4 <- data.frame(A  = rnorm(5,0,0.08), B = rnorm(5,0,0.08))
xlimits <- c(-1,1)

t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t3Plot <- ggplot(t3)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t4Plot <- ggplot(t4)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

grid.arrange(t1Plot,t2Plot, t3Plot, t4Plot, ncol=2)
```
\normalsize

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "100%"}
s <- function (table){ mean(table$A) - mean(table$B)}
whole <- rbind(t1,t2,t3,t4)
rawsd <- sd(c(whole$A,whole$B))
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
```
\normalsize
\pause

\vspace{-2mm}

- Raw sd in data is `r round(rawsd,3)`
- The sd of means is `r round(factor,3)`
- The WEAT score is `r round(numerator/factor,3)`   
- The largest effect size reported by @Caliskan2017semanticsBiases is 1.81!

## The problem with pre-averaging on realistic set-up

### Simulation with realistic set-up (16 predicates)

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "70%"}
set.seed(123456)
t1 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t2 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t3 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t4 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t5 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t6 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t7 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t8 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t9 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t10 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t11 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t12 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t13 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t14 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t15 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))
t16 <- data.frame(A  = rnorm(8,0,0.08), B = rnorm(8,0,0.08))

xlimits <- c(-1,1)

t1Plot <- ggplot(t1)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t1 in X")+xlim(xlimits)

t2Plot <- ggplot(t2)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t2 in X")+xlim(xlimits)

t8Plot <- ggplot(t8)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t3 in Y")+xlim(xlimits)

t9Plot <- ggplot(t9)+geom_point(aes(y = 0, x = A), size = 1, alpha = 0.5)+geom_point(aes(y = 0.1, x = B), col = "skyblue", size = 1, alpha = 0.5)+theme_tufte()+xlab("similarity")+scale_y_continuous(limits = c(-.05,.15),breaks = c(0,0.1), labels = c("A","B"))+ylab("group") + geom_vline(aes(xintercept = mean(A)), size = 0.3) + geom_vline(aes(xintercept = mean(B)), size = 0.3, col = "skyblue") +ggtitle("Similarities for term t4 in Y")+xlim(xlimits)

grid.arrange(t1Plot,t2Plot, t8Plot, t9Plot, ncol=2)

whole <- rbind(t1,t2,t3,t4)
rawsd <- sd(c(whole$A,whole$B))
factor <- sd(c(s(t1),s(t2),s(t3),s(t4)))
numerator <-  mean(s(t1),s(t2)) - mean(s(t3),s(t4))
```

\vspace{-2mm}

 Raw sd: `r round(rawsd,3)`, sd of means: `r round(factor,3)`, WEAT:  `r round(numerator/factor,3)`.

## The problem with pre-averaging on realistic set-up

### 10k simulations (same parameters)

\vspace{1mm}
\footnotesize
```{r,echo=FALSE,eval=TRUE,fig.align = "center",cache=TRUE, fig.show = "hold", out.width = "80%"}

set.seed(123)
biasesD0 <- numeric(10000)

>>>>

---
title: '\Large Taking uncertainty seriously  \newline \normalsize A  Bayesian approach
  to  word embedding bias estimation '
author: Alicja Dobrzeniecka \& Rafal  Urbaniak \footnotesize \newline (LoPSE research
  group, University of  Gdansk)
date: "Boston, April Fools' Day"
output:
  beamer_presentation:
    theme: Rafal_beamerSly1
    keep_tex: yes
    slide_level: 2
  slidy_presentation: default
fontsize: 10pt
classoption: x11names, dvipsnames, bibspacing, natbib, table
urlcolor: blue
bibliography: ../../references/cosineReferences1.bib
csl: ../../references/apa-6th-edition.csl
link-citations: yes
---

```{r setup, include=FALSE}
library(tidyverse)
library(gridExtra)
library(kableExtra)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(ggforce)
library(ggExtra)
library(ggthemes)
library(latex2exp)
th <- theme_tufte(base_size = 20)
options(kableExtra.latex.load_packages = FALSE)
```

@@

### Nearest words

\begin{itemize}
\item \textbf{philosophy}: philosophies, credo, ethos, principles, ethic, tenets, mantra, ideology, mindset, worldview

\item \textbf{sandwich}: sandwiches, burger, chicken sandwich, cheeseburger, burrito, burgers, pizza, turkey sandwich, hamburger, burritos
\end{itemize}

@@

### Some similarities from \emph{philosophy}

Logic (.47), Nietzsche (.32), Hegel (.32), analytic (.13), burger (.08), continental (.04), Russell (.04)
